Variance reduction for MCMC methods via martingale representations

### 2020-02-17 aap/Revision/vr-mcmc-revision-new.tex
update vr-mcmc-revision-new.tex

@@

dX_t=b(X_t)\, \rmd t+\sigma(X_t)\, \rmd W_t,\quad t\geq 0,	

@@

Lg=b^\top \nabla g+\frac{1}{2}\text{Tr}(\sigma^\top D^2g\sigma)

@@

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration

@@

\begin{thm}\label{prop:29032018a1}
 Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some kernels $(\Phi_l)_{l\geq 1}.$
Then,
for $p>j$, the following representation holds in \(L^2(P)\)
\begin{eqnarray}
\label{eq:mart_repr}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X_{l-1})\phi_k\left(\xi_{l}\right),
\end{eqnarray}
where
\begin{eqnarray}
\label{eq:coeff_mart}
a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
\end{eqnarray}
\end{thm}
\begin{cor}
If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form
\begin{eqnarray*}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X_{l-1})\phi_k\left(\xi_{l}\right)
\end{eqnarray*}
with
\begin{eqnarray*}
\bar a_{r,k}(x)=\mathsf{E}\left[\left.f(X_{r})\phi_k\left(\xi_{1}\right)\right|X_{0}=x\right], \quad r, k\in \mathbb{N}.
\end{eqnarray*}

@@

Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$

@@

and $\mathcal{G}_0=\mathrm{triv}$.

>>>>

\newcommand{\tcr}[1]{\textcolor{red}{#1}}

@@

\rmd X_t=b(X_t)\, \rmd t+\sigma(X_t)\, \rmd W_t,\quad t\geq 0,	

@@

Lg=b^\top \nabla g+\frac{1}{2}\operatorname{Tr}(\sigma^\top D^2g\sigma)
\eric{do we need a trace here ?}

@@

$\phi_k(\xi)$, $k\ge1$, are centered. As an example, we can take  multivariate Hermite polynomials for the ULA algorithm and a tensor product of Shifted Legendre polynomials for "uniform part" and Hermite polynomials for "Gaussian part"  of the random variable $\xi = (u, z)^T$ in MALA, as the Shifted Legendre polynomials are orthogonal with respect to the Lebesgue measure on \([0,1].\) In the sequel, we denote by $(\mathcal{G}_p)_{p\in \mathbb{Z}_+}$  the filtration

@@

\erici{It would be perhaps interesting to state the conditions satisfied by the functions $(\Phi_l)_{l \geq 1}$ here}
\begin{thm}\label{prop:29032018a1}
 Let $f$ be a Borel function $\mathbb{R}^{d}\to\mathbb R$ such that
$\mathsf{E}\left[\left|f(X_{p})\right|^{2}\right]<\infty$,   where $(X_{p})_{p\geq 0}$ is a Markov chain of the form \eqref{eq:chain_gen} for some sequence of functions $(\Phi_l)_{l\geq 1}.$
Then,
for $p>j$, the following representation holds in \(L^2(P)\)
\label{eq:mart_repr}
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}a_{p,l,k}(X_{l-1})\phi_k\left(\xi_{l}\right),
\label{eq:coeff_mart}
a_{p,l,k}(x)=\mathsf{E}\left[\left.f(X_{p})\phi_k\left(\xi_{l}\right)\right|X_{l-1}=x\right], \quad p\geq l, \quad k\in \mathbb{N}.
\begin{cor}
If  all the kernels $\Phi_l,$ $l\geq 1,$ are equal,  then  the representation \eqref{eq:mart_repr} takes the form
f(X_{p})=\mathsf{E}\left[\left.f(X_{p})\right|\mathcal G_{j}\right]+\sum_{k=1}^{\infty}\sum_{l=j+1}^{p}\bar a_{p-l,k}(X_{l-1})\phi_k\left(\xi_{l}\right)
\bar a_{r,k}(x)=\mathsf{E}\left[\left.f(X_{r})\phi_k\left(\xi_{1}\right)\right|X_{0}=x\right], \quad r, k\in \mathbb{N}.

@@

\tcr{Recall that $\mathcal{G}_l=\sigma(\xi_{1},\ldots,\xi_{l}),$

@@

and $\mathcal{G}_0=\mathrm{triv}$.}
\eric{do we need this, this is recalled ten lines above}
